Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-09-22 revision/r2/response_r2.tex

\externaldocument{forecastflows_r2}
\newcommand{\lref}[1]{}
% (\href{file:forecastflows_r1\#lintarget:SOMELABEL}{line \ref*{lin:SOMELABEL}})

\DeclareEmphSequence{\itshape\color{darkgrey}}
\usepackage[most]{tcolorbox}
% \newtcolorbox{mybox}[1][]{%
%     blanker, 
%     left skip=0em,
%     left=1em,
%     borderline west={1pt}{2pt}{darkgrey},
%     breakable,
%     #1}
\newtcolorbox{mybox}[1][]{%
    blanker, 
    left skip=0em,
    left=0em,
    borderline west={0pt}{0pt}{white},
    breakable,
    #1}

\begin{document}
\setlength{\parindent}{0cm}
\setlength{\parskip}{7pt}

Editor and reviewer comments are in \emph{gray italics} (we provide below the full context of the reviewers 1-2 comments below, and extract the parts of reviewer 3' comments that required changes to this manuscript), while our responses are in regular text. \\

@@

{\bf Response to editor's comments:}

We are happy to hear that the three reviewers recognized the value of our work and that the manuscript is ready for publication. We have worked to fix the last small points raised by the reviewers, and we would like to thank them again for their careful reading. Furthermore, we also particularly appreciate the perspective on the policy implications from M. Authier, and will reach out to him to discuss further directly (as both he and we appear to agree that this is perhaps the topic for another paper).

@@

\begin{mybox}
\emph{Line 215: comma after parameters}
\end{mybox}

Added.

\begin{mybox}
\emph{A couple typos in Box e.g., "calibrating submodes separately allows to
avoid" missing "us"?}
\end{mybox}

% emwSep22: I changed this in the text and here; your previous version needs an 'it' (makes it possible) but I wanted to remove 'fact' so adjusted more. 
Changed to: ``Calibrating submodels separately makes it easier to avoid the likely reality that [...]''

@@

\begin{mybox}
\emph{\textbf{Supplement Minor comments:}\\
Although the supplement works through an example far simpler than the
types of research questions discussed in the main text, it serves as a
reasonable (and accessible) example of how the workflow could be
implemented. There is a brief discussion towards the end of the
supplement about the ``scalability" of the proposed workflow, but it
leaves something to be desired. Have the authors applied this workflow
(or pieces of it) to examples that are more complex that could be cited,
or is that future work?}
\end{mybox}

... % emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers .... I might just say: We disagree that if the model is not sufficiently complex it leaves `something to be desired.' As we want this workflow to be approachable and leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), we have left it as is.

@@

\begin{mybox}
\emph{Step 2: so assuming populations are exchangeable? Is this reasonable?}
\end{mybox}

We have edited the text here to clarify this assumption:
\begin{quote}
``In this first hierarchical model, we thus assume that populations are exchangeable---an assumption we might refine later."
\end{quote}

\begin{mybox}
\emph{Step 2: priors on sigma\_alpha,sp and sigma\_beta\_sp do not seem
appropriate, these parameters have positive support, so should these be
half-normals (folded normals)?}
\end{mybox}

... % emwSep22: Just add a comment to the Stan code or text regarding how Stan works, and say: we understand this confusion and have updated the text.

@@

\begin{mybox}
\emph{Stan model not included so could not be assessed (model1\_nc2.stan), I
recommend sharing for transparency.}
\end{mybox}

Apologies that we forgot to share the Stan code. We have now added the code in the workflow example.

@@

\begin{mybox}
\emph{Figure 15, unclear what the posterior predictive distribution is of,
this detail could be added for transparency. Could also add many
posterior predictive datasets to the same plot and then overlay the
observed data to facilitate easier comparisons between y-tilde and y-obs.}
\end{mybox}

% Hmm I don't understand?
% emwSep22:  I think this must be the figure on pg 15 -- we show just one draw of the PPC (red), we could overlap more (like 10-30) in a lighter shade. And put the observed data black ON TOP so easier to see. I think this would be good to do.... just one draw of a PPC is okay, but not greatI IMHO.

{\bf Reviewer 3} % emwSep22:  I think you should extract only the parts that require a change to the current text. I will read the below once you have that done and am happy to read a separate letter to MA.

### 2025-09-22 revision/r2/forecastflows_r2.tex
Revisions to letter (have not reviewed main text) and started a README

\llabel{robust} This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately makes possible to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.  %emw24Jul: nice stuff here.

>>>>

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above). Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory. The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately makes it easier to avoid the likely reality that, if the submodels were fitted together (as a whole), many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.  %emw24Jul: nice stuff here.
